\section{Existential lower bounds}\label{sec:random}
Before we embark on our quest to prove lower bounds for interesting
families of polynomials, it is natural to ask as to what bounds
one can hope to achieve.  For a multivariate polynomial 
$f(\vecx) \in \FF[\vecx]$, denote by $S(f)$ the size of 
the smallest arithmetic circuit computing
$f$.  
\begin{theorem}
	{\bf [Folklore.]}
  For ``most'' polynomials $f(\vecx) \in \FF[\vecx]$ of degree $d$ 
  on $n$ variables we have 
  $$ S(f) \spaced{\geq} \Omega \inparen{\sqrt{\binom{n + d}{d}}}. $$
\end{theorem}
\begin{proof-sketch}
  We prove this here only in the situation where the underlying field
  $\FF$ is a finite field and refer the reader to another survey
  (\cite{ckw11}, Chapter 4) for a proof in the general case. So let
  $\FF = \FF_{q}$ be a finite field. Any line of a straight line
  program computing $f$ can be expressed as taking the product of two
  $\F_q$-linear combinations of previously computed values. Hence the
  total number of straight-line programs of length $s$ is at most
  $q^{O(s^2)}$. On the other hand there are $q^{\binom{n+d}{d}}$
  polynomials of degree $d$ on $n$ variables.  Hence most $n$-variate
  polynomials of degree $d$ require straight-line programs of length
  at least (equivalently arithmetic circuits of size at least) $s =
  \Omega \inparen{\sqrt{\binom{n+d}{d}}}$.
	\end{proof-sketch}
	
\noindent	Hrubes and Yehudayoff \cite {hy11} showed that in fact 
	most $n$-variate polynomials of degree $d$ {\em with zero-one 
	coefficients} have complexity at least $\Omega\inparen{\sqrt{\binom{n+d}{d}}}$. 
	Now it turns out that this is in fact a lower bound on the 
	number of multiplications in any circuit computing a random 
	polynomial. Lovett \cite{lovett11} complements this 
	nicely by giving a matching upper bound. Specifically, 
	it was shown in \cite{lovett11} that for any polynomial $f$ 
	of degree $d$ on $n$ variables there exists a circuit computing 
	$f$ having at most $\inparen{\sqrt{\binom{n + d}{d}}} \cdot (nd)^{O(1)} $ 
	multiplications.



%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "lowerbounds"
%%% End: 
