\chapter{Lower bounds for multi-$k$-ic models}\label{chap:multi-k-ic}

Recall that a multilinear depth three circuit is a sum of \emph{multilinear terms} that are polynomials of the form $T = \ell_1 \dots \ell_d$ where every variable $x \in X$ is present in at most one $\ell_i$. 
Kayal and Saha \cite{ks15} study more general depth three circuits that are sums of terms where very variable occurs in ``few'' linear factors. 

\begin{definition}[Multi-$k$-ic depth three circuits]
  A product of linear polynomials $T = \ell_1 \dots \ell_d$ is said to be a \emph{multi-$k$-ic term} if every variable $x \in X$ occurs in at most $k$ of the linear polynomials. 
A depth three circuit is said to be a \emph{multi-$k$-ic depth three circuit} if it is a sum of multi-$k$-ic terms. 
\end{definition} 

For example, the circuit computing $(x_1 + x_2)(x_2 + x_3)(x_1 +x_3) + (x_1 + x_3)(x_2 + 3x_3)$ multi-$2$-ic circuit. \\

Kayal and Saha \cite{ks15} studied the question of proving lower bound for such multi-$k$-ic depth-three circuits for small $k$ and they showed that the techniques of \cite{raz2004} can be generalized to give exponential lower bounds for multi-$k$-ic depth-three circuits for small $k$. 

\begin{theorem}[\cite{ks15}]\label{thm:multi-k-ic} There is an explicit $n$-variate polynomial $f \in \VNP$ such that any multi-$k$-ic depth-three circuit computing it must have size $2^{\Omega(n/2^{100k})}$. 
\end{theorem}

\section{Revisiting the measure}

In all the multilinear lower bounds we saw, the measure used was the rank of the partial derivative matrix under a random partition. 
We shall use the same measure, but keeping in mind that monomials could be non-multilinear. 
Once again, for a partition $X = Y \sqcup Z$, we shall define the matrix $M_{Y,Z}(f)$ to be the matrix whose rows are indexed by all (possibly non-multilinear) monomials $m_i$ in the $Y$ variables, and columns are indexed by all (possibly non-multilinear) monomials $m_j$ in the $Z$ variables with the entry being the coefficient of $m_i m_j$ in $f$. 
We shall abuse notation and use $\CM{Raz}_{Y,Z}(f)$ to refer to the rank of $M_{Y,Z}(f)$. 

As in the earlier lower bounds, we shall take a random partition $X = Y \sqcup Z$ and study $\CM{Raz}_{Y,Z}(f)$ for a small multi-$k$-ic circuit. 

\begin{remark*}\rm 
We shall take a slight deviation from the way we chose partitions earlier. 
Here, we shall take every variable $x\in X$ and map it to $Y$ or $Z$ with equal probability. 
Thereby, it is possible that $|Y|$ is not exactly $|X|/2$ but it would nevertheless be close to it with high probability. 
This modification gives us the useful feature that the random partition is made of independent events. 
This would turn out to be useful in the following calculations. 
\end{remark*}

\section{Proof of Theorem~\ref{thm:multi-k-ic}}

The proof strategy will be the same as earlier. 
For a random polynomial with degree in every variable bounded by $k$, we expect $\CM{Raz}_{Y,Z}(f)$ to be $\Omega(k^{\min(|Y|,|Z|)})$. 
We shall show that for a multi-$k$-ic term $T = \ell_1\dots \ell_d$, under a random partition $X = Y \sqcup Z$, the measure $\CM{Raz}(T)$ will be far from $k^{n/2}$ with high probability, if $n = |X|$. 

The proof that we shall describe here is a simplification of the original proof of \cite{ks15}. 
This proof would be very similar to the proof of \autoref{lem:raz-depth-three}. 

\begin{lemma}\label{lem:multi-k-ic-ub}
Let $X = Y \sqcup Z$ be a random partition. 
If $T$ is a multi-$k$-ic term, then with probability $\inparen{1 - \exp(-\frac{|X|}{2^{6k+1}})}$
\[
\CM{Raz}_{Y,Z}(f) \spaced{\leq} (1+k)^{\min(|Y|,|Z|)}\cdot \exp\inparen{-\frac{|Y|}{2^{3k+1}}}
\]
\end{lemma}


Let us quickly recall how the proof of \autoref{lem:raz-depth-three} proceeded. 
We essentially showed that for any linear polynomial involving ``few'' variables, with some non-trivial probability the random partition would map all the variables to one side of the partition. 
Thus, if there are many such linear polynomials, the is a large fraction of the linear polynomials that do not contribute to $\CM{Raz}_{Y,Z}(T)$. 
If there are very few such ``small'' linear polynomials, then the degree would have cannot be too large and we can use a trivial bound. 
The proof here shall proceed in a very similar fashion. 


\begin{proofof}{\autoref{lem:multi-k-ic-ub}}
Let $T = \ell_1 \dots \ell_d$. 
We shall call a factor $\ell$ to be ``large'' if the number of variables it depends on is at least $3k$, and let $T_L$ be the product of all $\ell_i$'s that are ``large'', and let $T_S$ be the product of the remaining $\ell_i$'s. 


\autoref{obs:pdm-multiplicativity}, generalized to this setting, gives that $\CM{Raz}_{Y,Z}(T) \leq \CM{Raz}_{Y,Z}(T_L) \cdot \CM{Raz}_{Y,Z}(T_S)$. 
Thus, it suffices to bound each term separately. 
The easy case is, as one would expect, handling $T_L$. 
Suppose $Y$ is the smaller of the sets $Y$ and $Z$. 
Let us list down every variable in $Y$ that occurs in $T_L$ in order as $y_1,\dots, y_{r_L}$ (with repetition). 
Let $r_L = (1-\delta) |Y|k$ for some $0\leq \delta \leq 1$. 
A trivial bound for $\CM{Raz}_{Y,Z}(T_L)$ is 
\begin{equation}\label{eqn:TL-bound}
\CM{Raz}_{Y,Z}(T_L)\spaced{\leq} 2^{\deg(T_L)} \spaced{\leq} 2^{r_L/3k} \spaced{\leq} 2^{(1-\delta)|Y|/3}
\end{equation}
The trickier case is with $T_S$ but intuitively the setting is very similar to what we encountered in the proof of \autoref{lem:raz-depth-three}. 
Since any factor $\ell$ of $T_S$ depends on at most $3k$ variables, with probability at least $\inparen{\frac{1}{2^{3k-1}}}$, all the variables of $\ell$ would be on the same side of the partition. 
Thus, on expectation, there would be at least $\inparen{\frac{\deg(T_S)}{2^{3k-1}}}$ factors that would not contribute to $\CM{Raz}_{Y,Z}(T_S)$ at all. 
In \autoref{lem:raz-depth-three}, we could show that the number of such factors is concentrated around the expectation since the linear polynomials were disjoint and the events were independent. 
However, in this setting a variable can occur in multiple factors. 
Nevertheless, one can still use the fact that every variable occurs in at most $k$ factors to establish a concentration very similar to Chernoff's Bounds. 
The following beautiful theorem of Gavinsky, Lovett, Saks and Srinivasan \cite{GLSS12} is exactly what we need. 

\begin{theorem}[\cite{GLSS12}]
Let $X_1,\dots, X_n$ be independent random variables. 
Let $E_1,\dots, E_r$ be boolean random variables that are functions of $\inbrace{X_i}$ such that each $X_i$ influences at most $k$ of the $E_j$'s. 
If $\Pr[E_i = 1] \geq p$, then for any $\epsilon > 0$, we have
\[
\Pr\insquare{E_1 + \dots + E_r  \spaced{\leq} (p - \epsilon) r } \spaced{\leq} e^{-2\epsilon^2 r/k}
\]
\end{theorem}

Again, let us list down every variable in $Y$ that occurs in $T_S$ in order as $y_1,\dots, y_{r_S}$ (listed with repetition). 
Note that $r_S + r_L \leq |Y|k$, and since $r_L = (1-\delta)k|Y|$,  we have that $r_S \leq \delta |Y|$. 
Let $E_i$ be the indicator random variable that is $1$ if all the variables of the factor  $\ell$ that contains $y_i$ are mapped to the same side of the partition; we shall call such an instance variable $y_i$ as \emph{ineffective}. 
In other words, the set of ineffective instances $y_i$ (with $E_i = 1$) are those that do not contribute to $\CM{Raz}_{Y,Z}(T_S)$. 

Since each $\ell$ depends on at most $3k$ variables, we have that $\Pr[Y_i = 1] \geq  p= \frac{1}{2^{3k-1}}$. 
Let us set $\epsilon = \frac{p}{2}$ and use the above theorem. 
Hence, the probability at least $\inparen{1 - \exp(-\frac{r_S}{k2^{6k}})}$, there are at least $\inparen{\frac{r_S}{2^{3k}}}$ ineffective instances. 

For every variable $x \in Y$, let $d_x$ be the number of occurrences of $x$ that is not ineffective. 
We know that $\sum_{x\in Y} d_x \leq r_S \cdot \inparen{1 - \frac{1}{2^{3k}}}$. 
On the other hand, $\prod_{x\in Y} (1 + d_x)$ is an upper-bound on the number of non-zero rows of $M_{Y,Z}(T_S)$. 
Hence,
\begin{eqnarray*}
\CM{Raz}_{Y,Z}(T_S) \spaced{\leq}\prod_{x\in Y} (1 + d_i) & \leq & \inparen{\frac{\sum_{x\in Y} (1 + d_x)}{|Y|}}^{|Y|}\\
 & \leq & \inparen{\frac{|Y| + r_S \inparen{1 - \frac{1}{2^{3k}}}}{|Y|}}^{|Y|}\\
 & \leq & \inparen{1 + \delta k\inparen{1 - \frac{1}{2^{3k}}}}^{|Y|}\\
 & \leq & \inparen{1+\delta k}^{|Y|} \cdot \inparen{1 - \frac{1}{2^{3k+1}}}^{|Y|}\\
 & \leq & \inparen{1+k}^{\delta |Y|}\exp\inparen{-\frac{|Y|}{2^{3k+1}}}
\end{eqnarray*}
Combining this with \eqref{eqn:TL-bound}, we get with probability at least $\inparen{1 - \exp(-\frac{|X|}{2^{6k+1}})}$
\[
\CM{Raz}_{Y,Z}(T) \spaced{\leq} (1+k)^{\min(|Y|,|Z|)}\cdot \exp\inparen{-\frac{|Y|}{2^{3k+1}}}
\]
\end{proofof}

\noindent
Using an union bound over all multi-$k$-ic terms, we obtain the simple corollary. 

\begin{corollary}\label{cor:multi-k-ic-ub}
Let $C = T_1 + \dots + T_s$ be a multi-$k$-ic circuit over variables $X$. 
Then for a random partition $X = Y \sqcup Z$, with probability at least $\inparen{1 - s  \cdot \exp(-\frac{|X|}{2^{6k+1}})}$, we have
\[
\CM{Raz}_{Y,Z}(C) \spaced{\leq} s \cdot (1+k)^{\min(|Y|,|Z|)}\cdot \exp\inparen{-\frac{|Y|}{2^{3k+1}}}
\]
In particular, if $s < \exp(\frac{|X|}{2^{6k+2}})$, then this happens with probability at least $1/2$. 
\end{corollary}

All that is left to do is find an explicit $f$ such that $\CM{Raz}_{Y,Z}(f) = (k+1)^{\min(|Y|,|Z|)}$ with high probability and we would have our lower bound. 
Here is one example that is a slight generalization of the construction in \autoref{sec:fullrankpoly}  of Raz and Yehudayoff~\cite{ry08} that is defined as follows. 
We shall again work over the field $\F(\setdef{\omega_{a,b,c},\omega_{a,b}}{a,b,c\leq 2n})$. 

\[
f_{i,j}^{(k)} =  \begin{cases}
 \sum\limits_{r=0}^k x_i^r x_{i+1}^r & \text{if $j = i=1$}\\
 0  & \text{if $j -i$ is even}\\
 \inparen{\sum\limits_{r=0}^k x_i^r x_{j}^r} \cdot f_{i+1,j-1}^{(k)} \cdot \omega{i,j}& \\
 \quad + \quad \sum\limits_{\ell=i+1}^{j-1} f_{i,\ell}^{(k)} \cdot f_{\ell+1,j}^{(k)} \cdot \omega_{i,\ell,j}\quad\quad & \text{otherwise} 
\end{cases}
\]
\noindent
The following lemma generalizes follows almost directly from \autoref{lem:fullrankpoly}. 

\begin{lemma}
The polynomial $f = f_{1,2n}^{(k)}$ defined above has the property that for every partition $X = \inbrace{x_1,\dots, x_{2n}} = Y \sqcup Z$, we have
\[
\CM{Raz}_{Y,Z}(f) \spaced{=} (k+1)^{\min(|Y|,|Z|)}
\]
Further, this polynomial can be computed by a linear sized arithmetic circuit and hence is in $\VP$. 
\end{lemma}

Combining this with \autoref{cor:multi-k-ic-ub} directly gives the lower bound of \autoref{thm:multi-k-ic} for a polynomial in $\VP$. 

\begin{remark}
  There has been a subsequent improvement by Kayal, Saha and Tavenas which I hope to add to this chapter soon. 
\end{remark}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "fancymain"
%%% End:
